---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE, warning = FALSE
)
```

# Nowcast right censored epidemological counts

[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R-CMD-check](https://github.com/epiforecasts/epinowcast/workflows/R-CMD-check/badge.svg)](https://github.com/epiforecasts/epinowcast/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/epiforecasts/epinowcast/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiforecasts/epinowcast)

[![Universe](https://epiforecasts.r-universe.dev/badges/epinowcast)](https://epiforecasts.r-universe.dev/)
[![MIT
license](https://img.shields.io/badge/License-MIT-blue.svg)](https://github.com/epiforecasts/epinowcast/blob/master/LICENSE.md/)
[![GitHub contributors](https://img.shields.io/github/contributors/epiforecasts/epinowcast)](https://github.com/epiforecasts/epinowcast/graphs/contributors) 

[![DOI](https://zenodo.org/badge/422611952.svg)](https://zenodo.org/badge/latestdoi/422611952)

This package contains tools to enable flexible and efficient hierarchical nowcasting of right censored epidemiological counts using a semi-mechanistic Bayesian method with support for both day of reference and day of report effects. 

## Installation

### Installing the package

Install the stable development version of the package with:

```{r, eval = FALSE}
install.packages("epinowcast", repos = "https://epiforecasts.r-universe.dev")
```

Install the unstable development from GitHub using the following, 

```{r, eval = FALSE}
remotes::install_github("epiforecasts/epinowcast", dependencies = TRUE)
```

### Installing CmdStan

If you don't already have CmdStan installed then, in addition to installing `epinowcast`, it is also necessary to install CmdStan using CmdStanR's 
`install_cmdstan()` function to enable model fitting in `epinowcast`. A suitable C++ toolchain is also required. Instructions are provided in the [_Getting started with
CmdStanR_](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) vignette. See the [CmdStanR documentation](https://mc-stan.org/cmdstanr/) for further details and support. 

```{r, eval = FALSE}
cmdstanr::install_cmdstan()
```

## Quick start

In this quick start we use COVID-19 hospitalisations by date of positive test in Germany available up to the 1st of October 2021 to demonstrate the specification and fitting of a simple nowcasting model using `{epinowcast}`.Examples using more complex models are available in the package vignettes and in the papers linked to in the literature vignette.

### Package

As well as `epinowcast` this quick start makes use of `data.table` and `ggplot2` which are both installed when `epinowcast` is installed.

```{r}
library(epinowcast)
library(data.table)
library(ggplot2)
```

### Data

These data are sourced from the [Robert Koch Institute via the Germany Nowcasting hub](https://github.com/KITmetricslab/hospitalization-nowcast-hub/wiki/Truth-data#role-an-definition-of-the-seven-day-hospitalization-incidence) where they are deconvolved from weekly data and days with negative reported hospitalisations are adjusted. Below we first filter for a snapshot of retrospective data available 30 days before the 1st of October that contains 30 days of data and then produce the nowcast target based on the latest available hospitialisations by date of positive test.

```{r}
nat_germany_hosp <- germany_covid19_hosp[location == "DE"][age_group %in% "00+"]
nat_germany_hosp[report_date <= as.Date("2021-10-01")]

retro_nat_germany <- enw_retrospective_data(
  nat_germany_hosp,
  rep_days = 30, ref_days = 30
)
retro_nat_germany
```

```{r}
latest_germany_hosp <- enw_latest_data(nat_germany_hosp, ref_window = c(60, 30))
latest_germany_hosp
```

### Data preprocessing and model specification

Process reported data into format required for `epinowcast` and return in a `{data.table}`. At this stage specify grouping (i.e age, location) if any. It can be useful to check this output before beginning to model to make sure everything is as expected.

```{r}
pobs <- enw_preprocess_data(retro_nat_germany, max_delay = 30)
pobs
```

Construct an intercept only model for the date of reference using the metadata produced by `enw_preprocess_data()`. Note that `{epinowcast}` uses a sparse design matrix to reduce runtimes so the design matrix shows only unique rows with `index` containing the mapping to the full design matrix.

```{r}
reference_effects <- enw_formula(pobs$metareference[[1]])
reference_effects
```

Construct a model with a random effect for the day of report using the metadata produced by `enw_preprocess_data()`.
```{r}
report_effects <- enw_formula(pobs$metareport[[1]], random = "day_of_week")
report_effects
```

### Model fitting

First compile the model. This step can be left to `epinowcast` but here we want to use multiple cores per chain to speed up model fitting and so need to compile the model with this feature turned on.
```{r}
model <- enw_model(threads = TRUE)
```

We now fit the model and produce a nowcast using this fit. Note that here we use two chains each using two threads as a demonstration but in general using 4 chains is recommended. Also note that here we have silenced fitting progress and potential warning messages for the purposes of keeping this quick start short but in general this should not be done.

```{r}
options(mc.cores = 2)
nowcast <- epinowcast(pobs,
  model = model,
  report_effects = report_effects,
  reference_effects = reference_effects,
  save_warmup = FALSE, pp = TRUE,
  chains = 2, threads_per_chain = 2,
  refresh = 0, show_messages = FALSE
)
```

### Results

Print the output from `{epinowcast}` which includes diagnostic information, the data used for fitting, and the `{cmdstanr`} object.

```{r}
nowcast
```

Summarise the nowcast for the latest snapshot of data.

```{r}
summary(nowcast, probs = c(0.05, 0.95))
```

Plot the summarised nowcast against currently observed data (or optionally more recent data for comparison purposes).

```{r nowcast}
plot(nowcast, obs = latest_germany_hosp)
```

Plot posterior predictions for observed notifications by date of
report as a check of how well the model reproduces the observed data.

```{r pp, fig.width = 16, fig.height = 16, message = FALSE}
plot(nowcast, type = "posterior") +
  facet_wrap(vars(reference_date), scale = "free")
```

## Citation

If using `epinowccast` in your work please consider citing it using the following,

```{r, echo = FALSE}
citation("epinowcast")
```

## How to make a bug report or feature request

Please briefly describe your problem and what output you expect in an [issue](https://github.com/epiforecasts/epinowcast/issues). If you have a question, please don't open an issue. Instead, ask on our [Q and A page](https://github.com/epiforecasts/epinowcast/discussions/categories/q-a).

## Contributing

We welcome contributions and new contributors! We particularly appreciate help on priority problems in the [issues](https://github.com/epiforecasts/epinowcast/issues). Please check and add to the issues, and/or add a [pull request](https://github.com/epiforecasts/epinowcast/pulls).

If interested in expanding the functionality of the underlying model note that `{epinowcast}` allows users to pass in their own models meaning that alternative parameterisations, for example altering the forecast model used for inferring expected observations, may be easily tested using the package infrastructure. Once this testing has been done alterations that increase the flexibility of the package model and improves its defaults are very welcome.

## Code of Conduct
  
Please note that the `forecast.vocs` project is released with a [Contributor Code of Conduct](https://epiforecasts.io/epinowcast/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
