---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Nowcast right censored epidemological counts

[![R-CMD-check](https://github.com/seabbs/epinowcast/workflows/R-CMD-check/badge.svg)](https://github.com/seabbs/epinowcast/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/seabbs/epinowcast/branch/main/graph/badge.svg)](https://app.codecov.io/gh/seabbs/epinowcast)

[![GitHub contributors](https://img.shields.io/github/contributors/seabbs/epinowcast)](https://github.com/seabbs/epinowcast/graphs/contributors) 

This package contains tools to enable flexible and efficient nowcasting of  right censored epidemiological counts using a semi-mechanistic method with adjustment available for both day of reference and day of report effects. 

## Installation


### Installing the package

Install the unstable development from GitHub using the following, 

```{r, eval = FALSE}
remotes::install_github("seabbs/epinowcast", dependencies = TRUE)
```

### Installing CmdStan

If you don't already have CmdStan installed then, in addition to installing `epinowcast`, it is also necessary to install CmdStan using CmdStanR's 
`install_cmdstan()` function to enable model fitting in `epinowcast`. A suitable C++ toolchain is also required. Instructions are provided in the [_Getting started with
CmdStanR_](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) vignette. See the [CmdStanR documentation](https://mc-stan.org/cmdstanr/) for further details and support. 

```{r, eval = FALSE}
cmdstanr::install_cmdstan()
```

## Quick start

## Citation

If using `epinowccast` in your work please consider citing it using the following,

```{r, echo = FALSE}
citation("epinowcast")
```

## How to make a bug report or feature request

Please briefly describe your problem and what output you expect in an [issue](https://github.com/seabbs/epinowcast/issues). If you have a question, please don't open an issue. Instead, ask on our [Q and A page](https://github.com/seabbs/epinowcast/discussions/categories/q-a).

## Contributing

We welcome contributions and new contributors! We particularly appreciate help on priority problems in the [issues](https://github.com/seabbs/epinowcast/issues). Please check and add to the issues, and/or add a [pull request](https://github.com/seabbs/epinowcast/pulls).

## Code of Conduct
  
Please note that the `forecast.vocs` project is released with a [Contributor Code of Conduct](samabbott.co.uk/epinowcast/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
