---
title: "ARIMA latent residuals: maths, priors, and usage"
description: "The maths behind the ARIMA(p, d, q) latent residuals in epinowcast, how to use them in any module's formula, and how to set their priors."
author: Epinowcast Team
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{ARIMA latent residuals: maths, priors, and usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

This vignette covers the maths behind `epinowcast`'s ARIMA(p, d, q) latent residual terms, how to use them in any module's formula, and how to set their priors.
It explains the kernel construction, the stationarity parameterisation, and the regularisation choices, with the trade-offs made explicit.
For worked usage and a comparison to `rw()` and lme4-style random effects, see the [latent processes vignette](latent-processes.html).
For the full model definition see the [model vignette](model.html).

# What an ARIMA term does

An `arima(time, by, p, d, q)` term in a formula declares a latent residual series that is added to the relevant linear predictor at every observation.
Writing the linear predictor of a module as $\eta_i$ for observation $i$ at time $t_i$ and group $g_i$:

$$\eta_i = X_i \beta + Z_i b + \epsilon_{t_i, g_i},$$

where $X_i \beta$ are the fixed effects, $Z_i b$ the random effects (including any `rw()` contributions, which are handled through the design matrix), and $\epsilon_{t_i, g_i}$ is the value of the ARIMA latent series for time $t_i$ and group $g_i$.
The residual evolves over the time axis declared in `arima(...)`; each observation just looks up the value at its own (time, group) index.

This is the same construction `brms` uses for `cor_arma`-style residual correlation on non-Gaussian likelihoods: a latent Gaussian residual added to the linear predictor, with ARMA structure imposed on its prior.

# Using ARIMA terms in a formula

Add a term to any module's formula with `arima()` or one of its aliases, which all build the same backend term:

- `rw(time)` — random walk, equal to `arima(time, p = 0, d = 1, q = 0)`.
- `ar(time, p)` — stationary autoregression, equal to `arima(time, p, d = 0, q = 0)`.
- `ma(time, q)` and `arma(time, p, q)` — moving-average and ARMA aliases.
- `arima(time, by, p, d, q)` — the general form.

`time` is the column the series evolves over, and the optional `by` gives each group its own shocks under shared dynamics.
The [latent processes vignette](latent-processes.html) works through these on the growth rate and compares them to `rw()` and lme4-style random effects; the [Priors](#priors) section below covers tuning their regularisation.

# The kernel decomposition

Let $T$ be the number of time points in the series and $G$ the number of groups.
The ARIMA latent is built from a $T \times G$ matrix of unit-normal shocks $z$ via

$$\epsilon = \sigma \, K(\phi, \theta, d) \, z,$$

where $\sigma$ is the latent standard deviation and $K$ is a $T \times T$ lower-triangular kernel that encodes the ARIMA structure.
The kernel factorises as

$$K(\phi, \theta, d) = D^d \, T(\psi),$$

where:

- $\psi \in \mathbb{R}^T$ is the truncated impulse response of the ARMA(p, q) process with autoregressive coefficients $\phi$ and moving-average coefficients $\theta$.
  It satisfies $\psi_1 = 1$ and the standard ARMA recursion [@boxjenkins]

  $$\psi_k = \sum_{j=1}^{\min(p, k-1)} \phi_j \psi_{k-j} + \mathbb{1}\{k - 1 \le q\} \, \theta_{k-1}, \quad k = 2, \dots, T.$$

- $T(\psi) \in \mathbb{R}^{T \times T}$ is the lower-triangular Toeplitz matrix with first column $\psi$.
  Multiplying $T(\psi) z$ applies the ARMA filter to the shocks: each entry $(T(\psi) z)_t = \sum_{k=1}^{t} \psi_{t-k+1} z_k$.

- $D \in \mathbb{R}^{T \times T}$ is the lower-triangular cumulative-sum operator, with ones on and below the diagonal.
  $D^d$ integrates $d$ times.
  $D^0 = I$ leaves the ARMA series unchanged; $D^1$ produces an integrated series of order one; higher $d$ generalise.

Two properties of this factorisation are useful:

- **$D^d$ is data-only.** It can be precomputed and does not contribute to the autodiff graph in Stan.
- **$T(\psi)$ is parameter-dependent.** It is rebuilt every gradient evaluation, but it is built once globally and applied to all shocks in a single matrix multiplication.

Composing the two and applying once to shocks gives the full ARIMA(p, d, q) latent:

```text
psi  <- arma_impulse(phi, theta, T)        # impulse response, length T
T_K  <- lower_toeplitz(psi)                # T x T ARMA kernel
D_d  <- cumulative_op(T, d)                # T x T integration operator
K    <- D_d * T_K                          # T x T composed kernel
eps  <- sigma * K * z                      # T x G latent
```

The Stan implementation of these steps lives in `inst/stan/functions/arima_kernel.stan`.

# Centring the integrated residual

When the residual is integrated ($d \ge 1$) the cumulative-sum operator $D^d$ gives $\epsilon$ a free overall level: the first shock propagates through the cumulative sum and shifts the whole path, so the level of $\epsilon$ trades off against the intercept $\alpha$ of the linear predictor.
This is a ridge in the joint posterior and slows sampling, the same way an un-centred design matrix correlates an intercept with its slopes.

When the module has a free intercept, each group's residual is mean-centred,

$$\tilde\epsilon_{\cdot, g} = \epsilon_{\cdot, g} - \frac{1}{T} \sum_{t=1}^{T} \epsilon_{t, g},$$

so the residual is a pure mean-zero deviation and the intercept owns the level.
To keep the prior on the original-scale intercept, the removed level is recovered into it,

$$\alpha = \alpha_c - \bar m, \qquad \bar m = \frac{1}{G} \sum_{g=1}^{G} \frac{1}{T} \sum_{t=1}^{T} \epsilon_{t, g},$$

where $\alpha_c$ is the sampled (centred) intercept and the user's prior is placed on $\alpha$.
The shift is additive, so the change of variables has unit Jacobian and the prior keeps its meaning.
For a single group this is an exact reparameterisation; for several groups the shared level is preserved while per-group deviations from the mean are re-expressed.
This mirrors how `brms` centres its population-level design matrix and how EpiNow2 centres the cumulative Gaussian process for the reproduction number, and on a weekly random-walk growth model it roughly halves the sampling time at the `adapt_delta` these models use.

# Stationarity via partial autocorrelations

The autoregressive part is only well behaved if it is *stationary*.
A stationary AR(p) process forgets its past, so its impulse response decays and the latent series stays bounded; a non-stationary one does the opposite, with an impulse response that grows without bound until the sampler diverges.
Putting a prior directly on $\phi$ is awkward because the stationary region of $\phi$-space is an irregular shape that is hard to sample inside.

We sidestep this with a standard reparameterisation.
Instead of $\phi$ we work with the *partial autocorrelations* $r_1, \dots, r_p$, each free to range over $(-1, 1)$.
Every such $r$ corresponds to exactly one stationary $\phi$ and vice versa, so sampling $r$ in the simple box $(-1, 1)^p$ stays in the stationary region by construction.
The map from $r$ to $\phi$ is the **Durbin-Levinson recursion** — the same recursion classically used to fit AR models from sample autocorrelations, run here in the forward direction [@barndorff].
For $k = 1, \dots, p$, starting from $\phi_1^{(1)} = r_1$,

$$\phi_k^{(k)} = r_k, \quad \phi_j^{(k)} = \phi_j^{(k-1)} - r_k \phi_{k-j}^{(k-1)}, \quad j = 1, \dots, k-1,$$

and the final coefficients are $\phi_j = \phi_j^{(p)}$.
By construction all roots of $1 - \phi_1 z - \dots - \phi_p z^p$ lie outside the unit circle, which is exactly the stationarity condition.
This is the same approach `brms` and `rstanarm` use for AR(p) priors.

In Stan we declare `vector<lower=-1, upper=1>[p] expr_arima_pacf` directly, then transform to $\phi$ via `pacf_to_phi(pacf)` inside the transformed parameters / model block.
The bounds put an implicit uniform prior on each $r_k$ over $(-1, 1)$; override it through the priors mechanism if needed.

MA invertibility is not enforced explicitly.
For typical values it is rarely a sampling problem; if it becomes one for a specific use case, the same partial-autocorrelation transform applies symmetrically on $\theta$.

The kernel produces $\epsilon \in \mathbb{R}^{T \times G}$, but observations live in a flat vector.
Each observation carries its own time and group index, so the contribution to the linear predictor is a single matrix lookup per observation — negligible next to building the kernel.

# Group structure: dependent vs independent

When `by` is supplied each group has its own column of unit-normal shocks `z[, g]`, but the AR/MA parameters $\phi$, $\theta$, and the latent standard deviation $\sigma$ are shared across groups.
The kernel $K$ is built once and applied to each column of the $T \times G$ shock matrix in one matrix multiplication, with each group's series being driven by its own innovations under a common dynamic.
Per-group parameters (each group with its own $\phi$, $\theta$, and $\sigma$) are a planned extension that would loop over $g$ building per-group $K_g$; until that lands the helpers do not expose a `type` argument.

# Special cases

The factorisation makes several familiar models special cases of the same code path:

- `arima(time, p = 0, d = 0, q = 0)` is degenerate and rejected by the constructor.
- `arima(time, p = 0, d = 1, q = 0)` reduces to a random walk: $\psi = (1, 0, 0, \dots)$, so $T(\psi) = I$ and $K = D$ is the cumulative-sum operator. **`rw(time)` is now a thin user-facing wrapper that constructs exactly this term**; both go through the same backend and produce identical posteriors. The pure-differencing case (no AR or MA structure) is detected in `arima_filter()` and short-circuited to a per-column `cumulative_sum()` so the kernel is never materialised — the cost is $O(T G)$ per evaluation, the same as the previous bespoke random-walk implementation.
- `arima(time, p = 1, d = 0, q = 0)` is a stationary AR(1): $\psi_k = \phi^{k-1}$, $T(\psi)$ is the standard AR(1) Toeplitz, $K = T(\psi)$.
- `arima(time, p = 0, d = 1, q = 1)` is an integrated MA(1) (the random-walk-with-correlated-shocks model commonly seen in epidemiology).

The `arima_kernel_matrix()` function is unit-tested against these cases against `stats::ARMAtoMA` and direct R reference recursions; see `tests/testthat/test-stan_arima_kernel.R`.

# Priors {#priors}

The default priors are deliberately weak.
The latent scale $\sigma$ and the partial autocorrelations $r_k$ can both be overridden through the standard `priors` argument of the relevant module; the shocks and the MA terms are fixed.

- $z_{t,g} \sim \mathcal{N}(0, 1)$ — fixed by the non-centred parameterisation; not user-tunable.
- $\sigma \sim \mathcal{N}_+(0, 0.2)$ (half-normal) — controls how strongly the latent series can deviate from zero. Set through the `<prefix>_arima_sigma` prior.
- $r_k \sim \mathrm{Uniform}(-1, 1)$ — the default, implicit from the parameter bounds. Set through the `<prefix>_arima_pacf` prior: a positive standard deviation switches to a $\mathcal{N}(\mu, s)$ prior truncated to $(-1, 1)$, shrinking the autocorrelations toward $\mu$. A standard deviation of zero keeps the uniform default.
- $\theta_l \sim \mathcal{N}(0, 1)$ when MA terms are present — weakly informative.

Here `<prefix>` is the module the term sits on (`expr` for the growth rate, `refp` for the parametric reference mean, and so on; see [Where you can use it](#where)).
For example, to shrink the growth-rate AR term toward weak autocorrelation rather than leaving it uniform:

```{r}
library(epinowcast)
pacf_prior <- data.frame(variable = "expr_arima_pacf", mean = 0, sd = 0.3)
# epinowcast(..., priors = pacf_prior)
```

The half-normal prior on $\sigma$ is the most consequential in practice; tighten it if the latent residual is competing with the rest of the linear predictor for variance.
A tighter `pacf` prior is the lever for discouraging the near-unit-root AR behaviour that the uniform default permits.

# Where you can use it {#where}

`arima()` and its aliases work on every module that takes a formula, each with its own prior prefix:

- `enw_expectation()` — the growth rate (`expr`) and the latent-to-obs proportion (`expl`).
- `enw_reference()` — the parametric mean (`refp`) and the non-parametric logit hazards (`refnp`).
- `enw_report()` — report-date logit hazards (`rep`).
- `enw_missing()` — the missing-reference proportion (`miss`).

Putting an `arima()` term on the reference delay mean, for instance, models a reporting delay that drifts over time, while an `arima()` on the growth rate models the epidemic trend itself.

# Limitations

- One ARIMA term per formula. The R side aborts if multiple are supplied; lifting this requires the Stan side to gain a vector of latent series rather than a single one.
- AR stationarity is enforced; MA invertibility is not (rarely a problem in practice).
- Grouped series share AR/MA parameters and latent standard deviation across groups; per-group parameters are a planned extension. Until that lands `arima()` and `rw()` do not take a `type` argument.
- The kernel is built dense ($T \times T$) every iteration. For very long series (thousands of time points) the autodiff cost grows as $O(T^2)$; in that regime a recursive form (computing $\epsilon$ via the ARMA filter directly) would be preferable.
- When the parametric reference standard deviation is modelled (`model_refp > 1`), it shares the ARIMA shocks and kernel of the mean but carries an independent scale (`refp_arima_sd_sigma`). The two residuals are therefore perfectly correlated up to that scale rather than driven by separate shock series. A separate sd process is a planned extension.

# References
