---
title: Discretised distributions
description: "Distributions and their discretisation in epinowcast"
author: Epinowcast Team
output: rmarkdown::html_document
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
vignette: >
  %\VignetteIndexEntry{Discretised distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette describes the parametric delay distributions that are currently available in `epinowcast` and explains how they are internally discretised.

```{r setup, include = FALSE}
has_deps <- requireNamespace("primarycensored", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  echo = TRUE, eval = has_deps, fig.width = 7, fig.height = 4
)
```

```{r libraries}
library(data.table)
library(ggplot2)
```

## Available distributions

The currently available parametric delay distributions are continuous probability distributions with (up to) two parameters $\mu_{g,t}$ and $\upsilon_{g,t}$. The table below provides a link to the definition of each distribution, specifies how the parameters $\mu_{g,t}$ and $\upsilon_{g,t}$ are mapped to the parameters of the distribution (according to the referenced definition), and states the resulting mean of the distribution (before discretization and adjustment for the assumed maximum delay).

| Distribution 	| Parametrization | Mean |
| :--------------:	| :------------:  | :------------------:	|
| [Log-normal](https://mc-stan.org/docs/functions-reference/lognormal.html)  | $\mu=\mu_{g,t}$, $\sigma = \upsilon_{g,t}$	| $\exp(\mu_{g,t}+\frac{\upsilon_{g,t}^2}{2})$ 	|
| [Exponential](https://mc-stan.org/docs/functions-reference/exponential-distribution.html)  | $\beta = \exp(-\mu_{g,t})$	| $\exp(\mu_{g,t})$ |
| [Gamma](https://mc-stan.org/docs/functions-reference/gamma-distribution.html)        |	$\alpha = \exp(\mu_{g,t})$, $\beta = \upsilon_{g,t}$ | $\exp(\mu_{g,t})/\upsilon_{g,t}$ |
| [Log-logistic](https://en.wikipedia.org/wiki/Log-logistic_distribution) |	$\alpha = \exp(\mu_{g,t})$, $\beta = \upsilon_{g,t}$ | $\frac{\exp(\mu_{g,t})\,\pi/\upsilon_{g,t}}{\sin(\pi/\upsilon_{g,t})}$ |

## Discretisation and adjustment for maximum delay

In `epinowcast`, delays are modelled in discrete time and with an assumed maximum delay (specified via the `max_delay` argument).
The continuous delay distributions must therefore be discretised and adjusted for the maximum delay.
`epinowcast` provides two discretisation methods, selected with the `discretisation` argument of `enw_reference()`.
The default (`"logit_hazard"`) uses a uniform-interval approximation.
The `"primarycensored"` method uses the double interval censoring discretisation from the [primarycensored](https://primarycensored.epinowcast.org) package [@primarycensored].

It is helpful to separate two distinct adjustments.
The first is discretisation: turning the continuous delay into a probability mass over integer delays $d = 0, 1, 2, \dots$, with each $p_d$ defined for an infinite maximum delay so that $\sum_{d=0}^{\infty} p_d = 1$.
The second is conditioning on the maximum delay $D$: restricting attention to delays $d \le D$ and renormalising so that the truncated probabilities sum to 1, i.e. $p^{\prime}_{d} = p_d / \sum_{j=0}^{D} p_j$.
The first step is about how a continuous distribution becomes discrete; the second is about right truncation at $D$.

### Default uniform-interval approximation (`"logit_hazard"`)

The exact form of the discretisation is complex due to the interaction between primary and secondary events.
Rather than modelling this explicitly, the default method approximates it by assuming a uniform censoring interval of 2 days for each delay.
This comes from assuming daily censoring of both the primary and secondary events, which together define the delay distribution, and ignoring potential interactions between primary and secondary events.
As a result, the probability of reporting a delay of $d$ days equals the probability of reporting a delay of $d+1$ days or less, minus the probability of reporting a delay of $d-1$ days or less.

Let $F^{\mu_{g,t}, \upsilon_{g,t}}$ be the cumulative distribution function of a continuous probability distribution of delays with parameters $\mu_{g,t}$ and $\upsilon_{g,t}$.
Before conditioning on the maximum delay, the discretised probability of a delay of $d$ days is
$$p_{g,t,d} = F^{\mu_{g,t}, \upsilon_{g,t}}(d+1) - F^{\mu_{g,t}, \upsilon_{g,t}}(d-1),$$
except for $d = 0$, where $p_{g,t,0} = F^{\mu_{g,t}, \upsilon_{g,t}}(1)$.

Conditioning on the maximum delay $D$ then divides by $F^{\mu_{g,t}, \upsilon_{g,t}}(D+1) + F^{\mu_{g,t}, \upsilon_{g,t}}(D)$, which is the total mass placed on delays $0, \dots, D$ by the overlapping intervals above.
This ensures that the conditioned probabilities sum to 1 and can be interpreted as conditioning the distribution on reporting before the maximum delay.

Because of the uniform-interval discretisation, the discrete delay distribution obtained this way only approximates the original double interval censored distribution, and the approximation is worse for shorter delays.

### Primary event censored discretisation (`"primarycensored"`)

The `"primarycensored"` method instead computes the double interval censored probability mass directly, accounting for the primary event window, the secondary (reporting) interval, and right truncation at $D$ [@primarycensored].
Writing the delay as the time between a primary event (e.g. infection), assumed uniform over a window of width 1, and a secondary event observed in a daily interval, the probability of a censored delay $d$ is obtained by integrating the continuous delay density over both the primary window and the secondary interval, then conditioning on the maximum delay $D$.
The Stan implementation is vendored from the [primarycensored](https://primarycensored.epinowcast.org) package and uses analytical solutions for the supported distributions (the exponential is handled as a gamma with shape one).
This avoids the uniform-interval approximation used by the default method and so recovers the double interval censored distribution more exactly, including at short delays.
It is currently available for the exponential, lognormal and gamma distributions.

The same `primarycensored` machinery underpins delay handling elsewhere in the ecosystem.
[`epidist`](https://epidist.epinowcast.org) estimates delay distributions from individual line-list data, and [`EpiNow2::estimate_dist()`](https://epiforecasts.io/EpiNow2/) fits them from aggregated count data; both are powered by `primarycensored`, with the main difference being the data structure they expect.
Estimating a delay with one of those tools and then passing it to `enw_reference()` with `discretisation = "primarycensored"` keeps the censoring assumptions consistent across the workflow.

To use it, pass `discretisation = "primarycensored"` to `enw_reference()`:

```r
reference <- enw_reference(
  ~1,
  distribution = "lognormal",
  discretisation = "primarycensored",
  data = pobs
)
```

### Recovering the double interval censored distribution

The `"primarycensored"` discretisation reproduces the reference probability mass function computed directly by `primarycensored::dprimarycensored()`.
The chunk below contrasts the two epinowcast discretisations against that reference for a lognormal delay (`meanlog = 1`, `sdlog = 0.5`) truncated at a maximum delay of 15 days.
The default uniform-interval approximation diverges from the reference at short delays, while the `"primarycensored"` method matches it.

```{r recovery}
dmax <- 15
meanlog <- 1
sdlog <- 0.5

# Reference double interval censored PMF from primarycensored: a uniform
# primary window (pwindow = 1) and daily secondary censoring (swindow = 1),
# right truncated at D = dmax.
reference <- primarycensored::dprimarycensored(
  0:(dmax - 1), plnorm, pwindow = 1, swindow = 1, D = dmax,
  meanlog = meanlog, sdlog = sdlog
)

# epinowcast default uniform-interval discretisation (overlapping 2-day
# intervals, normalised to sum to 1), matching discretised_logit_hazard().
cdf <- plnorm(0:dmax, meanlog, sdlog)
default <- numeric(dmax)
default[1] <- cdf[2]
if (dmax > 1) default[2] <- cdf[3]
if (dmax > 2) default[3:dmax] <- cdf[4:(dmax + 1)] - cdf[1:(dmax - 2)]
default <- default / sum(default)

pmf <- data.table(
  delay = rep(0:(dmax - 1), 2),
  probability = c(default, reference),
  method = rep(c("logit_hazard (default)", "primarycensored"), each = dmax)
)
print(pmf[delay <= 3])
```

The largest discrepancy is at the shortest delays, exactly where the uniform-interval approximation is least accurate:

```{r recovery-plot}
ggplot(pmf, aes(x = delay, y = probability, colour = method)) +
  geom_point() +
  geom_line() +
  labs(
    x = "Delay (days)", y = "Probability",
    colour = "Discretisation",
    title = "Discretised lognormal delay (meanlog = 1, sdlog = 0.5)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The `"primarycensored"` curve is the double interval censored target; the default curve is an approximation to it.


