---
title: "Latent process options for the growth-rate model"
description: "Random walks, group random slopes, and ARIMA(p, d, q) residuals — the time-series structures available in the formula interface for the growth rate."
author: Epinowcast Team
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: library.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-numeric-superscript-brackets.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Latent process options for the growth-rate model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

The expectation module's `r` formula controls the latent log growth rate.
Several options exist for giving it time-series structure; each is a different choice for how `r` is allowed to vary over time and across groups.

This vignette walks through three of them on the same small slice of the German COVID-19 hospitalisation data:

- **Random walk on weeks** via `rw(week)`. The growth rate has iid Gaussian increments per week; the cumulative sum of those increments is what enters the linear predictor.
- **Independent per-week-and-group random effects** via `(1 | week:.group)`. Each (week, group) combination has its own random level drawn from a shared Gaussian, with no time-correlation structure across weeks.
- **ARIMA(p, d, q) residuals** via `arima(day, age_group, p, d, q)`. The growth rate is built up through a parameter-dependent kernel applied to unit-normal shocks, supporting arbitrary autoregressive and moving-average orders and integration.

The fits use `enw_pathfinder()` for fast variational inference so the vignette stays snappy at build time.
For a real analysis switch to NUTS via `enw_sample()`.

# Setup

```{r setup, message = FALSE}
library(epinowcast)
library(data.table)
library(ggplot2) # nolint: unused_import_linter. Required by plot.epinowcast().
library(knitr)
```

```{r}
options(mc.cores = 2)
```

We work with three German age strata reported nationally over five weeks of reference dates ending 28 days before the latest available reports.
The same `pobs` is used for every fit so any difference between the resulting nowcasts is down to the latent-process choice rather than the data.

```{r data}
nat_germany_hosp <- germany_covid19_hosp[
  location == "DE" & age_group %in% c("00+", "00-04", "80+")
]

retro <- nat_germany_hosp |>
  enw_filter_report_dates(remove_days = 28) |>
  enw_filter_reference_dates(include_days = 35)

pobs <- suppressWarnings(
  enw_preprocess_data(retro, by = "age_group", max_delay = 14)
)
```

A common fit configuration is reused across the models.
`enw_pathfinder()` is a fast approximate sampler that suits feature tours and exploration; for production fits use `enw_sample()` instead.

```{r fit-opts}
fit <- enw_fit_opts(
  sampler = enw_pathfinder,
  num_paths = 4, single_path_draws = 50, draws = 200,
  refresh = 0, show_messages = FALSE
)
```

# Random walk on weeks

`rw(week)` adds one Gaussian increment per week; the cumulative sum of those increments is added to `r`.
Mathematically this is equivalent to a (correlated) Gaussian prior over the per-week growth rates with a single shared standard deviation.

```{r rw-fit, results = "hide"}
nowcast_rw <- epinowcast(
  pobs,
  expectation = enw_expectation(r = ~ rw(week), data = pobs),
  obs = enw_obs(family = "negbin", data = pobs),
  fit = fit
)
```

```{r rw-plot, fig.cap = "Nowcast under a random walk on weeks."}
plot(nowcast_rw)
```

# Independent per-(week, group) random effects

`(1 | week:.group)` adds a separate random level for every (week, group) cell, drawn from a shared Gaussian.
Unlike the random walk, there is no time correlation across weeks; each week starts afresh.
This is the natural choice when weekly fluctuations look like noise rather than drift.

```{r re-fit, results = "hide"}
nowcast_re <- epinowcast(
  pobs,
  expectation = enw_expectation(
    r = ~ 1 + (1 | week:.group), data = pobs
  ),
  obs = enw_obs(family = "negbin", data = pobs),
  fit = fit
)
```

```{r re-plot, fig.cap = "Nowcast under independent per-(week, group) random effects."}
plot(nowcast_re)
```

# ARIMA(p, d, q) residuals

`arima(time, by, p, d, q, type)` adds a latent ARIMA(p, d, q) residual series to the linear predictor.
Differencing is applied via the cumulative-sum operator and the ARMA part via a parameter-dependent Toeplitz kernel built from the impulse response; both compose with a single matrix multiplication onto unit-normal shocks.
Setting `type = "dependent"` shares `phi`, `theta`, and `sigma` across groups (a single ARIMA process applied independently to each group's shocks), which is usually a sensible default unless groups are believed to differ in their dynamics.

`arima(time, d = 1, p = 0, q = 0)` recovers the same model as `rw(time)`; setting `p > 0` or `q > 0` adds autoregressive or moving-average structure on top.

```{r arima-fit, results = "hide"}
nowcast_arima <- epinowcast(
  pobs,
  expectation = enw_expectation(
    r = ~ 1 + arima(day, age_group, p = 1, d = 1, type = "dependent"),
    data = pobs
  ),
  obs = enw_obs(family = "negbin", data = pobs),
  fit = fit
)
```

```{r arima-plot, fig.cap = "Nowcast under an ARIMA(1, 1, 0) residual on day."}
plot(nowcast_arima)
```

The ARIMA-specific posterior summaries:

```{r arima-pars}
arima_pars <- summary(nowcast_arima, type = "fit")[
  grepl("expr_arima_(pacf|sigma)", variable),
  .(variable, mean, q5, q95)
]
kable(arima_pars, digits = 3)
```

# When to reach for which

| Latent option | Formula example | Useful when |
|---|---|---|
| Random walk | `rw(week)` | The growth rate evolves smoothly week-to-week with no preferred direction; the standard non-stationary smoother. |
| Independent random effects | `(1 \| week:.group)` | Weekly fluctuations look like noise around a stable mean rather than a drifting trend; no time correlation is imposed. |
| ARIMA | `arima(day, by, p, d, q)` | You want explicit autocorrelated short-range structure on the residuals, e.g. `arima(p = 1, d = 0)` for stationary AR(1) deviations around a smooth trend, or `arima(p = 1, d = 1)` for AR(1) on the differences. |

The three options can be combined within the same formula; for example `~ 1 + week + arima(day, p = 1)` gives a linear time trend with AR(1) residuals.
The `by` argument on `arima()` and `rw()` lets you fit independent series per group, with `type = "dependent"` sharing the standard deviation and dynamics across groups.
