---
title: "Visualising Preprocessed Data"
description: "Understanding reporting patterns before model fitting."
author: Epinowcast Team
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Visualising Preprocessed Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Before fitting a nowcasting model it is useful to explore the reporting data to understand the delay structure, identify anomalies, and check that preprocessing has worked as expected.
This vignette walks through the plot types available for `enw_preprocess_data` objects.

# Setup

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  dpi = 120,
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages, message = FALSE}
library(epinowcast)
library(data.table)
library(ggplot2)
```

# Data

We use the COVID-19 hospitalisation data included in the package, filtered to national-level counts in Germany.
We create a retrospective snapshot as if we were standing on 22 August 2021 with 40 days of data and a maximum reporting delay of 40 days.

```{r data}
nat_germany_hosp <-
  germany_covid19_hosp[location == "DE"][age_group == "00+"] |>
  enw_filter_report_dates(latest_date = "2021-10-01")

retro_nat_germany <- nat_germany_hosp |>
  enw_filter_report_dates(remove_days = 40) |>
  enw_filter_reference_dates(include_days = 40)
```

# Preprocessing

```{r preprocess}
pobs <- enw_preprocess_data(retro_nat_germany, max_delay = 40)
pobs
```

The preprocessed object bundles several data tables together.
We can now visualise these data using the `plot` method, which dispatches to specialised plot functions depending on the `type` argument.

# Latest observations

The default plot type (`"obs"`) shows the latest cumulative case counts by reference date.
This is the data the model will attempt to nowcast.

```{r obs, fig.cap = "Latest reported hospitalisations by date of positive test."}
plot(pobs, type = "obs")
```

# Cumulative reporting delay

The `"emp_rep_cum"` type shows the cumulative fraction of cases reported by each delay group over time.
Reference dates where a large fraction is reported quickly appear as ribbons that reach the top of the plot early.
Dates where reporting is slow show wider gaps between ribbons.

```{r emp-rep-cum, fig.cap = "Cumulative fraction reported by delay group."}
plot(pobs, type = "emp_rep_cum")
```

When no `delay_group_thresh` is supplied the thresholds are generated automatically from `max_delay`.
Custom thresholds can highlight specific delay windows of interest.

```{r emp-rep-cum-custom, fig.cap = "Cumulative reporting with custom delay thresholds."}
plot(
  pobs, type = "emp_rep_cum",
  delay_group_thresh = c(0, 1, 3, 7, 14, 41)
)
```

# Reporting delay heatmap

The `"emp_rep_frac"` type shows the fraction of cases reported in each delay group as a tile plot.
This is useful for spotting systematic shifts in reporting speed over time and day-of-the-week effects.

```{r emp-rep-frac, fig.cap = "Fraction of cases reported by delay group and reference date."}
plot(pobs, type = "emp_rep_frac")
```

# Reporting delay quantiles

The `"emp_rep_quant"` type plots empirical quantiles of the reporting delay distribution for each reference date.
By default the 10th, 50th, and 90th percentiles are shown.
Stable lines indicate a stationary delay distribution; trends suggest the delay structure is changing.

```{r emp-rep-quant, fig.cap = "Empirical delay quantiles over time."}
plot(pobs, type = "emp_rep_quant")
```

Custom quantiles can be specified.

```{r emp-rep-quant-custom, fig.cap = "Median and interquartile range of reporting delays."}
plot(pobs, type = "emp_rep_quant", quantiles = c(0.25, 0.5, 0.75))
```

# Notifications by delay group

The `"emp_ts_del"` type produces a stacked bar plot showing the number of notifications by reference date, coloured by how long they took to be reported.
This gives a combined view of the volume of reports and their timeliness.

```{r emp-ts-del, fig.cap = "Notifications by reference date coloured by reporting delay."}
plot(pobs, type = "emp_ts_del")
```

# Using the individual plot functions

Each plot type corresponds to an exported function that can be called directly for more control.

| Plot type        | Function                     |
|:-----------------|:-----------------------------|
| `"obs"`          | `enw_plot_obs()`             |
| `"emp_rep_cum"`  | `enw_plot_emprep_cum()`      |
| `"emp_rep_frac"` | `enw_plot_emprep_frac()`     |
| `"emp_rep_quant"`| `enw_plot_emprep_quant()`    |
| `"emp_ts_del"`   | `enw_plot_emp_ts_del()`      |

These return standard `ggplot2` objects so layers, facets, and themes can be added freely.

```{r custom-plot}
enw_plot_emprep_frac(
  pobs, delay_group_thresh = c(0, 1, 3, 7, 14, 41)
) +
  scale_fill_viridis_c() +
  ggtitle("Reporting delay heatmap with viridis scale")
```

# Helper functions

Two helper functions underpin the delay-based plots and can be used independently for custom analyses.

`enw_cat_new_confirm()` categorises notifications into delay groups and computes empirical reporting proportions.

```{r cat-new-confirm}
nc <- enw_cat_new_confirm(
  pobs, delay_group_thresh = c(0, 1, 3, 7, 14, 41)
)
head(nc)
```

`enw_emp_quant_by_reference()` computes empirical delay quantiles by reference date.

```{r emp-quant}
eq <- enw_emp_quant_by_reference(pobs)
head(eq)
```
