---
title: "Visualising Preprocessed Data"
description: "Understanding reporting patterns before model fitting."
author: Epinowcast Team
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Visualising Preprocessed Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Before fitting a nowcasting model it is useful to explore the reporting data to understand the delay structure, identify anomalies, and check that preprocessing has worked as expected.
This vignette walks through the plot types available for `enw_preprocess_data` objects.

# Setup

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  dpi = 120,
  collapse = TRUE,
  comment = "#>"
)
```

```{r packages, message = FALSE}
library(epinowcast)
library(data.table)
library(ggplot2)
```

# Data

We use the COVID-19 hospitalisation data included in the package, filtered to national-level counts in Germany.
We create a retrospective snapshot as if we were standing on 1 October 2021 with 80 days of reference dates and a maximum reporting delay of 40 days.

```{r data}
nat_germany_hosp <-
  germany_covid19_hosp[location == "DE"][age_group == "00+"] |>
  enw_filter_report_dates(latest_date = "2021-10-01")

retro_nat_germany <- nat_germany_hosp |>
  enw_filter_report_dates(remove_days = 40) |>
  enw_filter_reference_dates(include_days = 80)
```

# Preprocessing

```{r preprocess}
pobs <- enw_preprocess_data(retro_nat_germany, max_delay = 40)
pobs
```

The preprocessed object bundles several data tables together.
We can now visualise these data using the `plot` method, which dispatches to specialised plot functions depending on the `type` argument.

All delay-based plots below are affected by right truncation: the most recent reference dates have not yet had enough time for all reports to arrive, so their delay distributions will appear to be shorter than the true distribution.
Keep this in mind when interpreting the rightmost portion of each plot.

# Latest observations

The default plot type (`"obs"`) shows the latest cumulative case counts by reference date.
This is the data the model will attempt to nowcast.

```{r obs, fig.cap = "Latest reported hospitalisations by date of positive test."}
plot(pobs, type = "obs")
```

The apparent drop in counts at the right edge of the series is a hallmark of right truncation: reports for recent reference dates have simply not had enough time to arrive.
This is precisely the signal that nowcasting aims to correct.

# Cumulative reporting delay

The `"delay_cumulative"` type shows the cumulative fraction of cases reported by each delay group over time.
Reference dates where a large fraction is reported quickly appear as ribbons that reach the top of the plot early.
Dates where reporting is slow show wider gaps between ribbons.

```{r emp-rep-cum, fig.cap = "Cumulative fraction reported by delay group."}
plot(pobs, type = "delay_cumulative")
```

When no `delay_group_thresh` is supplied the thresholds are generated automatically from `max_delay`.
Custom thresholds can highlight specific delay windows of interest.

```{r emp-rep-cum-custom, fig.cap = "Cumulative reporting with custom delay thresholds."}
plot(
  pobs, type = "delay_cumulative",
  delay_group_thresh = c(0, 1, 3, 7, 14, 41)
)
```

If the ribbons are stable across reference dates the delay distribution is roughly stationary, which may justify a simpler model without time-varying delay components.
Drift or shifts in the ribbons indicate the delay structure is changing and should be modelled.

# Reporting delay heatmap

The `"delay_fraction"` type shows the fraction of cases reported in each delay group as a tile plot.

```{r emp-rep-frac, fig.cap = "Fraction of cases reported by delay group and reference date."}
plot(pobs, type = "delay_fraction")
```

Here we can see changes in colour across columns that line up with day of the week, which indicates the delay distribution depends on the reference weekday.
The cumulative plot shows how fast reports accumulate overall, while the heatmap isolates where within the delay distribution a change is happening.
If the cumulative plot is stable but the heatmap shows shifting bands, reports are arriving at the same aggregate pace but being redistributed across delays.
Conversely, stable heatmap bands combined with a changing cumulative plot point to a shift in total reporting speed rather than in the shape of the delay distribution.

# Reporting delay quantiles

The `"delay_quantiles"` type plots empirical quantiles of the reporting delay distribution for each reference date.
By default the 10th, 50th, and 90th percentiles are shown.

```{r emp-rep-quant, fig.cap = "Empirical delay quantiles over time."}
plot(pobs, type = "delay_quantiles")
```

Lower quantiles (e.g. the 10th percentile) are less affected by right truncation because early reports have had time to arrive.
Higher quantiles (e.g. the 90th percentile) are more heavily truncated because they depend on late-arriving reports that may not yet have been observed for recent reference dates.
A sudden drop in the higher quantiles at the right edge is therefore expected and does not necessarily indicate a real change in reporting speed.

Quantile lines summarise the delay distribution as a single number per reference date, which makes small temporal trends easier to read than from the heatmap but hides the full shape.
Use the quantile plot to check whether the median and tails drift over time; fall back to the heatmap when you need to see which delays are responsible for a change.

Custom quantiles can be specified.

```{r emp-rep-quant-custom, fig.cap = "Median and interquartile range of reporting delays."}
plot(pobs, type = "delay_quantiles", quantiles = c(0.25, 0.5, 0.75))
```

# Notifications by delay group

The `"delay_counts"` type produces a stacked bar plot showing the number of notifications by reference date, coloured by how long they took to be reported.
This combines the volume of reports with their timeliness in a single view.

```{r emp-ts-del, fig.cap = "Notifications by reference date coloured by reporting delay."}
plot(pobs, type = "delay_counts")
```

Compared to the cumulative and heatmap plots, which show proportions, this plot puts absolute counts on the y-axis.
Use it when you care about the size of each delay group in context with the overall reporting volume, for example when deciding whether a noisy-looking right edge is supported by many notifications or by only a handful.

# Using the individual plot functions

Each plot type corresponds to an exported function that can be called directly for more control.

| Plot type             | Function                        |
|:----------------------|:--------------------------------|
| `"obs"`               | `enw_plot_obs()`                |
| `"delay_cumulative"`  | `enw_plot_delay_cumulative()`   |
| `"delay_fraction"`    | `enw_plot_delay_fraction()`     |
| `"delay_quantiles"`   | `enw_plot_delay_quantiles()`    |
| `"delay_counts"`      | `enw_plot_delay_counts()`       |

These return standard `ggplot2` objects so layers, facets, and themes can be added freely.
Grouped data are auto-faceted by `.group`; pass `facet = FALSE` to disable and supply your own layout.

```{r custom-plot}
enw_plot_delay_fraction(
  pobs, delay_group_thresh = c(0, 1, 3, 7, 14, 41)
) +
  scale_fill_viridis_c() +
  ggtitle("Reporting delay heatmap with viridis scale")
```

# Helper functions

Two helper functions underpin the delay-based plots and can be used independently for custom analyses.

`enw_delay_categories()` categorises notifications into delay groups and computes empirical reporting proportions.

```{r cat-new-confirm}
nc <- enw_delay_categories(
  pobs, delay_group_thresh = c(0, 1, 3, 7, 14, 41)
)
head(nc)
```

`enw_delay_quantiles()` computes empirical delay quantiles by reference date.

```{r emp-quant}
eq <- enw_delay_quantiles(pobs)
head(eq)
```
