---
title: "simple-rt-estimation"
description: ""
author: Sam Abbott
opengraph:
  image: 
    src: figures/performance-1.png
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simple-rt-estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(epinowcast)
```

- Introduce the package and sign post to further documentation
- Introduuce the problem targeted in this vignette
- Introduce a synthetic data example 
- Introduce the model mathematically and with code (i.e. chunk by chunk).
- Fit the model to the synthetic data
- Visualise the nowcast, the Rt estimate, posterior predictions + compare to actual synthetic values.
- Introduce the case study data and additional assumptions required.
- Fit the same model to case study data
- Show nowcast, Rt estimate, posterior predictions, and delay estimate vs retrospective delay.
- Discuss results, highlight limitations, and sign post to further documentation.